---
title: "Heart Disease"
author: "Sejal-Shannon"
date: "11/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chunk 1: #Libraries 
```{r}
library(plotly)
library(stats)
library(tidyr)
library(ggplot2)
library(dplyr)
library(data.table)
library(tidyverse)
library(pastecs)
library(dbplyr)
library(Hmisc)
library(corrplot)
library(pscl)
library(AER)
library(robust)
library(ROCit)
```

#Chunk 2: Working directory 
```{r}
getwd()
```

#Chunk 3: Read in the data 
```{r}
heart_dataframe <- read.csv("./data/heart.csv")
head(heart_dataframe)
str(heart_dataframe)
```


#Chunk 4: Check NAs and for select variable check for missing/invalid data 
Missing data checked for variables Age, RestingBP, Cholesterol and MaxHR, as these are the variables where a value zero is not valid. Missing/Zero values replaced by NA
```{r}
colSums(is.na(heart_dataframe))

invalid_Age <- heart_dataframe$Age == 0
heart_dataframe$Age[invalid_Age] <- NA

invalid_RestingBP<- heart_dataframe$RestingBP == 0
heart_dataframe$RestingBP[invalid_RestingBP] <- 132.5

#As variable cholesterol has large number(172) of invalid values. Replace zero values with random values between 1st Quartile and mean. 
invalid_Cholesterol <- heart_dataframe$Cholesterol == 0
heart_dataframe$Cholesterol[invalid_Cholesterol]<- runif(172, min = 173.2, max = 198.8)

invalid_MaxHR <- heart_dataframe$MaxHR == 0
heart_dataframe$MaxHR[invalid_MaxHR] <- NA

nrow(heart_dataframe)
#na.omit(heart_dataframe)
colSums(is.na(heart_dataframe))
summary(heart_dataframe)
table(heart_dataframe$Cholesterol)
head(heart_dataframe, 50)
```

#Chunk 5: DataType Conversion. Applicable variables to factors 
```{r}
#sex
heart_dataframe$Sex[heart_dataframe$Sex == "F"] <- "Female"
heart_dataframe$Sex[heart_dataframe$Sex == "M"] <- "Male"
heart_dataframe$Sex <- as.factor(heart_dataframe$Sex)


#ChestPainType
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "TA"] <- "Typical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ATA"] <- "Atypical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "NAP"] <- "Non-anginal pain"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ASY"] <- "Asymptomatic angina"
heart_dataframe$ChestPainType <- as.factor(heart_dataframe$ChestPainType)


#FastingBS 
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 0] <- "Low  (<120 mg/dl)"
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 1] <- "High (>120 mg/dl)"
heart_dataframe$FastingBS <- as.factor(heart_dataframe$FastingBS)

#RestingECG
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "Normal"] <- "0 (normal)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "ST"] <- "1 (ST)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "LVH"] <- "2 (LVH)"
heart_dataframe$RestingECG <- as.factor(heart_dataframe$RestingECG)

#Exercise induced angina
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "N"] <- "No"
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "Y"] <- "Yes"
heart_dataframe$ExerciseAngina <- as.factor(heart_dataframe$ExerciseAngina)

#ST Slope
heart_dataframe$ST_Slope <- as.factor(heart_dataframe$ST_Slop)

#Heart Disease
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "0"] <- "Health"
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "1"] <- "Heart Disease"
heart_dataframe$HeartDisease <- as.factor(heart_dataframe$HeartDisease)

head(heart_dataframe$HeartDisease,20)
summary(heart_dataframe)
str(heart_dataframe)
```


#Chunk 6: Characterizing distributions of NUMERIC variables and correcting for skew
```{r}
#Histograms for continuous integer/dbl variables 

#AGE
age_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Age, fill = heart_dataframe$Age, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "Chocolate")+
  xlab("Age in years")+
  ylab("frequency")+
  ggtitle("Distribution of Age in database")
age_distribution 

qqplot_Age<- qplot(sample = heart_dataframe$Age)+
xlab("Age in years")+
ylab("frequency")
qqplot_Age
#qqplot o/p indicates age of subjects are normally distributed.


#CHOLESTEROL 
cholesterol_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Cholesterol, fill = heart_dataframe$Cholesterol, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "azure2")+
  xlab("Serum Cholesterol Levels in mg/dl")+
  ylab("frequency")+
  ggtitle("Distribution of Serum Cholesterol Levels across database")
cholesterol_distribution 

qqplot_cholesterol<- qplot(sample = heart_dataframe$Cholesterol, na.rm = TRUE)+
xlab("Serum Cholesterol Levels in mg/dl")+
ylab("frequency")
qqplot_cholesterol

describe(heart_dataframe$Cholesterol)
stat.desc(heart_dataframe$Cholesterol, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$Cholesterol)
#Based on qqplot, descriptive statistics and histogram, Cholesterol Levels appears to be NOT normally distributed, has a positive skew. Kurtosis of 4.6 (>3) indicates data is leptokurtic.

#log transformation to correct positive skew 
log_cholesterol <- log(heart_dataframe$Cholesterol)
log_RBP_hist <- hist(log_cholesterol, col = "azure2", border = "black", main = "Distribution of Serum Cholesterol Levels(transformed)", xlab = "Serum Cholesterol Levels in mg/dl", ylab = "frequency")

#RESTING BLOOD PRESSURE 
RBP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingBP, fill = heart_dataframe$RestingBP, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "brown")+
  xlab("Blood Pressure is mmHg")+
  ylab("frequency")+
  ggtitle("Distribution of Resting Blood Pressure Readings")
RBP_distribution 

describe(heart_dataframe$RestingBP)
stat.desc(heart_dataframe$RestingBP, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$RestingBP)

#log transformation to correct skew
log_RBP <- log(heart_dataframe$RestingBP)
log_RBP_hist <- hist(log_RBP, col = "brown", border = "black", main = "Distribution of Resting Blood Pressure Readings(transformed)", xlab = "Resting Blood Pressure in mmHg", ylab = "frequency")

#MAXIMUM HEART RATE 
MHR_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$MaxHR, fill = heart_dataframe$MaxHR))+
  geom_histogram(color = "Black", fill = "darkorchid1")+
  xlab("Maximum Heart Rate in bpm")+
  ylab("frequency")+
  ggtitle("Distribution of Maximum Heart Rate across database")
MHR_distribution 

#Distribution appears to be nearly normally distributed. 
describe(heart_dataframe$MaxHR)
stat.desc(heart_dataframe$MaxHR, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$MaxHR)



#OLD PEAK 
Oldpeak_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Oldpeak, fill = heart_dataframe$Oldpeak))+
  geom_histogram(color = "Black", fill = "cadetblue")+
  xlab("Oldpeak - ST depression induced by exercise relative to rest")+
  ylab("frequency")+
  ggtitle("Distribution of ST depression induced by exercise relative to rest")
Oldpeak_distribution #Positively skewed data 

qqplot_oldpeak<- qplot(sample = heart_dataframe$Oldpeak, na.rm = TRUE)+
xlab("Oldpeak - ST depression induced by exercise relative to rest")+
ylab("frequency")
qqplot_oldpeak

describe(heart_dataframe$Oldpeak)
stat.desc(heart_dataframe$Oldpeak, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$Oldpeak)  #confirms data is not normally distributed. 

#log to correct for severe positive skew 
log_oldpeak <- log(heart_dataframe$Oldpeak)
log_oldpeak
log_oldpeak_hist <- hist(log_oldpeak, col = "cadetblue", border = "black", main = "Distribution of ST depression induced by exercise relative to rest(transformed)", xlab = "Oldpeak - ST depression induced by exercise relative to rest", ylab = "frequency")


```

#Chunk 7: Characterizing distributions of FACTOR variables   
Bar plots used for factor variables and Histograms for continuous variables. 
```{r}
#SEX
gender_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Sex, fill = heart_dataframe$Sex))+
  geom_bar()+
  xlab("Gender")+
  ylab("frequency")+
  ggtitle("Distribution of Gender")+
  scale_fill_discrete(name = "Gender", labels = c("Female","Male"))
gender_distribution

#Data set contains significantly higher number of males(700) compared to females (200)

#CHEST PAIN TYPES 
CP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ChestPainType, fill = heart_dataframe$ChestPainType))+
  geom_bar()+
  xlab("Types of Chest Pain")+
  ylab("frequency")+
  ggtitle("Distribution of Types of Chest Pain")+
  scale_fill_discrete(name = "Types of Chest Pain", labels = c("Asymptomatic", "Atypica", "Non-Anginal", "Typical"))
CP_distribution

# percent distribution of chest pain types 
cp_table <- table(heart_dataframe$ChestPainType)
percent_cp <- round(cp_table/sum(cp_table) *100)
labels1 <- paste(names(cp_table), percent_cp)
labels2 <- paste(labels1, "%", sep = "")
pie(cp_table, labels = labels2, col = c("azure1", "azure2", "azure3", "azure4"), main = "Distribution of types of chest pain within heart dataset" ) 

#Atypical Angina is the most observed 

#FASTING BLOOD SUGAR LEVLS
FBS_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$FastingBS, fill = heart_dataframe$FastingBS))+
  geom_bar()+
  xlab("Fasting Blood Sugar Levels in mg/dl")+
  ylab("frequency")+
  ggtitle("Distribution of Fasting Blood Sugar Levels")+
  scale_fill_discrete(name = "Blood Sugar Levels", labels = c(">120 mg/dl","<120 mg/dl"))
FBS_distribution

#RESTING ECG
ECG_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingECG, fill = heart_dataframe$RestingECG))+
  geom_bar()+
  xlab("Resting ECG Results")+
  ylab("frequency")+
  ggtitle("Distribution of Resting ECG Results")+
  scale_fill_discrete(name = "ECG Results", labels = c("Normal", "ST", "LVH"))
ECG_distribution


#EXERCISE INDUCED ANGINA
Exangina_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ExerciseAngina, fill = heart_dataframe$ExerciseAngina))+
  geom_bar()+
  xlab("presence or absence of exercise induced angina ")+
  ylab("frequency")+
  ggtitle("Presence or Absence of Exercise Induced Angina")+
  scale_fill_discrete(name = "Exercise Induced Angina", labels = c("Absence","Presence"))
Exangina_distribution

#ST Slope
ST_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ST_Slope, fill = heart_dataframe$ST_Slope))+
  geom_bar()+
  xlab("ST Slope nature")+
  ylab("frequency")+
  ggtitle("Distribution of nature of the ST Slopes")+
  scale_fill_discrete(name = "ST Slopes", labels = c("Down", "Flat", "Up"))
ST_distribution

#HEART DISEASE
HD_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$HeartDisease, fill = heart_dataframe$HeartDisease))+
  geom_bar()+
  xlab("Heart Disease")+
  ylab("frequency")+
  ggtitle("Presence and absence of heart disease")+
  scale_fill_discrete(name = "Heart Disease", labels = c("Absence", "presence"))
HD_distribution
prop.table(table(heart_dataframe$HeartDisease)) 

#Within the data set ~55.3% have heart disease and ~44.7% do not.
```



#Chunk 8: Data conversion to all numeric (except HeartDisease) to obtain a complete correlation matrix and correlation plot 
```{r}
heart_dfnum <- subset(heart_dataframe, select = -c(HeartDisease))

#YN_HeartDisease, YN_HeartDisease_bin
#ST Slope
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "2"] <- "0"
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "1"] <- "1"
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "3"] <- "2"
heart_dfnum$ST_Slope <- as.numeric(heart_dfnum$ST_Slope)


#sex
heart_dfnum$Sex[heart_dfnum$Sex == "M"] <- "0"
heart_dfnum$Sex[heart_dfnum$Sex == "F"] <- "1"
heart_dfnum$Sex <- as.numeric (heart_dfnum$Sex)


#ChestPainType
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "TA"] <- "0"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "ATA"] <- "1"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "NAP"] <- "2"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "ASY"] <- "3"
heart_dfnum$ChestPainType <- as.numeric(heart_dfnum$ChestPainType)


heart_dfnum$FastingBS <- as.numeric(heart_dfnum$FastingBS)

#RestingECG
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "Normal"] <- "0"
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "ST"] <- "1"
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "LVH"] <- "2"
heart_dfnum$RestingECG <- as.numeric(heart_dfnum$RestingECG)

#Exercise induced angina
heart_dfnum$ExerciseAngina[heart_dfnum$ExerciseAngina == "N"] <-"0"
heart_dfnum$ExerciseAngina[heart_dfnum$ExerciseAngina == "Y"] <- "1"
heart_dfnum$ExerciseAngina <- as.numeric(heart_dfnum$ExerciseAngina)


log_dataframe <- data.frame(heart_dataframe$Age, log_cholesterol,log_RBP, log_oldpeak, heart_dataframe$MaxHR, heart_dfnum$ST_Slope, heart_dfnum$Sex, heart_dfnum$ChestPainType, heart_dfnum$FastingBS,heart_dfnum$RestingECG, heart_dfnum$ExerciseAngina)
str(log_dataframe)
colnames(log_dataframe) <- c("Age", "Cholesterol", "Resting Blood Pressure", "Old Peak", "Max Heart Rate" , "ST Slope", "Gender","Types of Chest Pain", "Fasting Blood Sugar Levels", "Resting ECG", "Exercise Induced Angina" )


head(heart_dfnum,20)
summary(heart_dfnum)
str(heart_dfnum)
```
 
#Chunk 9: Correlation Matrix and Plots 
```{r}
#originally numeric variables only 
heart_numeric_log <- select_if(log_dataframe, is.numeric)
heart_numeric_log
heart_numcor_log <- cor(heart_numeric, use = "complete", method = "spearman")
heart_numcor_log
corrplot(heart_numcor_log, method = "circle")

#all variables to numeric (except Outcome Variable) 
heart_allnum_cor <- cor(heart_dfnum, use = "complete.obs")
heart_allnum_cor 
heart_allnum_corplot <- corrplot(heart_allnum_cor, method = "circle", order = "alphabet", tl.cex = 0.7)

#All - transformed 
heart_logcor <- cor(log_dataframe, use = "complete.obs", method = "spearman")
heart_logcor
heart_logcorplot <- corrplot(heart_logcor, method = "square", order = "alphabet", tl.cex = 0.7)

```


#Chunk 10: Correlation graphs: Age and Lab Values VS Heart Disease 
```{r}
#Between Heart Disease and Age 
age_HD <- ggplot(heart_dataframe, aes(x = Age, fill = HeartDisease, color = HeartDisease))+
          geom_histogram(binwidth = 1, color = "black") +
          facet_wrap(vars(HeartDisease))+
          xlab("Age in Years")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Distribution of Heart Disease in relation to Age")
age_HD
#Histogram plots for presence Vs absence(Health) of heart disease have different distributions. 
#1. The distribution of presence of heart disease is negatively skewed.
#2. The distribution of absence of heart disease appears to be normally distributed. 
#3. Hence we can conclude that heart disease is more prevalent in older population compared to younger. 

#Between Resting blood pressure and presence/absence of heart disease
RBP_HD <- ggplot(heart_dataframe, aes( x = RestingBP, fill = HeartDisease))+
          geom_histogram()+
          xlab("Resting Blood Pressure in mmhg")+
          ylab(" Heart Disease")
# There does not appear to be a very strong relationship between resting BP and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 
RBP_HD

CHL_HD <- ggplot(heart_dataframe, aes( x = Cholesterol,  fill = HeartDisease))+
          geom_histogram()+
          xlab("Serum Cholesterol Levels in mg/dl")+
          ylab("Heart Disease")
CHL_HD
#Similarly, there does not appear to be a very strong relationship between Serum Cholesterol and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 

#Relationship between max HR achieved during exercise and presence and absence of heart disease 
MHR_HD <- ggplot(heart_dataframe, aes( x = MaxHR, fill = HeartDisease))+
          geom_histogram()+
          xlab("Maximum HR during exercise in bpm")+
          ylab("Heart Disease")
MHR_HD
#Max HR is the HR per minute reached during strenuous exercise. 
#There is a strong relationship between Max HR induced during exercise and presence and absence of heart disease. Subjects without heart disease are able to achieve a higher Max HR overall compared to subjects with heart disease. 

#Between Fasting blood sugar levels and presence/absence of heart disease
FBS_HD <- ggplot(heart_dataframe, aes(x = FastingBS, fill = HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Fasting Blood Sugar levles in mgdl")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Relationship between Fasting Blood Sugar Levels and Heart Disease")
FBS_HD
#Barplot above shows comparison between fasting blood sugar levels and presence or absence of heart disease. 
# Proportions in the plot indicate that there exists a relationship between fasting blood sugar levels and heart disease. 
# Fasting blood sugar levels of >120mg/dl (high) have increased rate of presence of heart disease compared to subjects with fasting blood sugar levels of <120mg/dl (low)

```


#Chunk 11: Correlation: Cardiovascular events and Heart Disease 
```{r}
#Between Exercise induced angina and presence and absence of heart disease
ExAng_HD <- FBS_HD <- ggplot(heart_dataframe, aes(x = ExerciseAngina, fill = HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Presence or Absence of exercise induced angina")+
          ylab("Heart Disease")+
          ggtitle("Relationship between Exercise Induced Angina and Heart Disease")
ExAng_HD
#Barplot above shows relationship between presence or absence of exercise induced angina Vs that of heart disease
# Proportions in the plot indicate that there exists a relationship between the two. 
# Rate of presence of exercise induced angina is significantly higher in subjects with heart disease compared to those with no heart disease (i.e. Health)

#Between ST segment depression induced by exercise and Heart Disease 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 

Oldpeak_HD <- ggplot(heart_dataframe, aes(x = Oldpeak, fill = HeartDisease, color = HeartDisease))+
          geom_histogram(binwidth = 0.5, color = "black") +
          facet_wrap(vars(HeartDisease))+
          xlab("ST segment depression values in mm")+
          ylab(" Heart Disease")+
          ggtitle("ST depression induced by exercise and presence and absence of heart disease")
Oldpeak_HD

#Histograms below show the relationship between distribution of ST segment depression induced by exercise and presence and absence of heart disease. 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 
#ST segment depression distribution in subjects without heart disease(Health) is positively skewed with majority subjects at 0.0 depression. 
#However, the distribution in subjects with heart disease shows significant number of subjects with ST segment depression >0.5 which is considered pathological. 

#Distribution of chest pain types in presence vs absence of heart disease

CPT_HD <- ggplot(heart_dataframe, aes( x = HeartDisease, fill = ChestPainType))+
  geom_bar(position = "dodge")+
  xlab(" Heart Disease")+
  ylab("Type of Chest Pain")+
  ggtitle("Distribution of chest pain types in presence vs absence of heart disease")
CPT_HD

```
#Chunk 12.0: Logistic Regression - PsurdoR initiation
```{r}
#pseudoR calculation function for later use 
logisticPseudoR2s <- function(LogModel) {
  dev <- LogModel$deviance 
  nullDev <- LogModel$null.deviance 
  modelN <-  length(LogModel$fitted.values)
  R.l <-  1 -  dev / nullDev
  R.cs <- 1- exp ( -(nullDev - dev) / modelN)
  R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
  cat("Pseudo R^2 for logistic regression\n")
  cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
  cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
  cat("Nagelkerke R^2           ", round(R.n, 3),    "\n")
}
```

#Chunk 12.1: Logistic Regression 
```{r}
summary(heart_dataframe)
table(heart_dataframe$HeartDisease)
prop.table(table(heart_dataframe$HeartDisease)) 
#55.34% have heart disease (508/918)
```
#Chunk 12.2 : Logistic Regression: Thresholding Outcome Variable HeartDisease and conversion to factor 
```{r}
#assigning threshold to Heart Disease 
heart_dataframe$YN_HeartDisease[heart_dataframe$HeartDisease == "Health"] <- "0"
heart_dataframe$YN_HeartDisease[heart_dataframe$HeartDisease == "Heart Disease"] <- "1"
table(heart_dataframe$YN_HeartDisease)
str(heart_dataframe$YN_HeartDisease)

heart_dataframe$YN_HeartDisease_bin <- as.factor(heart_dataframe$YN_HeartDisease)

heart_dataframe$YN_HeartDisease <- factor(heart_dataframe$YN_HeartDisease, levels = c(0,1), labels = c("Absence", "Presence"))

table(heart_dataframe$YN_HeartDisease)
str(heart_dataframe)
```

#Chunk 12.3: Logistic Regression - Building Model 1 = model1_full
```{r}
model1_full <- glm(YN_HeartDisease ~ Age + Sex + ChestPainType + RestingBP + Cholesterol+ FastingBS+RestingECG+ MaxHR+ ExerciseAngina+Oldpeak+ ST_Slope, data= heart_dataframe, family=binomial())
summary(model1_full)
```

#Chunk 12.4: Logistic Regression - Model 2(with significant variables only )
```{r}
#model 2 - no interactions
model2.0_reduced <- glm(YN_HeartDisease ~ Sex + ChestPainType + FastingBS + ExerciseAngina+Oldpeak+ ST_Slope,  data = heart_dataframe, family=binomial())
model2.0_reduced

coef(model2.0_reduced)
exp(coef(model2.0_reduced))
#Odds of Heart Disease Increases by a factor of 2.72  with every event of presence of Exercise Angina while holding rest constant.  
#Odds of Heart Disease Increased by a factor of 3.55 for every unit increase in ST_Slope(Flat) ??? while holding rest constant 
#Odds of Heart Disease increases by a factor of 1.45 for every unit of depression in ST segment induced by exercise (Oldpeak) 
#Odds of Heart Disease increases by a factor of 4.99 for subject being a Male while holding the rest constant. 



#model 2 
model2_reduced <- glm(YN_HeartDisease ~ Sex + ChestPainType + FastingBS + ExerciseAngina+Oldpeak+ ST_Slope + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heart_dataframe, family=binomial())
model2_reduced
summary(model2_reduced) 

coef(model2_reduced)
exp(coef(model2_reduced))

 

```
# Chunk 12.5: Removing Outliers for Model 3 
```{r}
#find outliers 
cooksd <- cooks.distance(model2_reduced)
cooksd
plot(cooksd, pch = "*", cex = 2) 

outlier <- c()

for (i in cooksd){
  if((is.na(i)== FALSE & i > (4/nrow(heart_dataframe))))
    {outlier <- append(outlier, i)
  }
}

length(outlier)
outlier_index <- which(cooksd %in% outlier)
outlier_index

heart_no_outlier <- heart_dataframe[-outlier_index,]
heart_no_outlier
nrow(heart_no_outlier)

```

#Chunk 12.6:  Logistic Regression Model 3: 
```{r}
heartdf_no_outlier <- data.frame(heart_no_outlier)
head(heartdf_no_outlier)
model3 <- glm(YN_HeartDisease ~ Sex + ChestPainType + FastingBS + ExerciseAngina+Oldpeak+ ST_Slope + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heartdf_no_outlier, family=binomial())
model2_reduced
summary(model3) 

coef(model3)
exp(coef(model3))
```

#Chunk 13: Comparing  model 1 & 2 performance - ANOVA 
```{r}
anova(model1_full, model2_reduced, test = "Chisq") 

#Model2_reduced has the significant p value p = 3.459e-05, suggests that model with significant predictors and their interactions fits better than model with all predictors. 
```

#Chunk 13.1: Comparing model2 performace - Pseudo R2 and adjusted R2 
```{r}
R_square <- 1- model2_reduced$deviance/model2_reduced$null.deviance
R_square

adjust_R_square <- - 1-(model2_reduced$deviance-(ncol(heart_dataframe)-1))/model2_reduced$null.deviance
adjust_R_square
```



