---
title: "Prediction of Heart Disease"
author: "Sejal Desai and Shannon Walsh"
date: "11/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Chunk 1: Libraries 
```{r}
library(plotly)
library(stats)
library(tidyr)
library(ggplot2)
library(dplyr)
library(data.table)
library(tidyverse)
library(pastecs)
library(dbplyr)
library(Hmisc)
library(corrplot)
library(pscl)
library(AER)
library(robust)
library(ROCit)
library(car)
library(pROC) 
library(caret)
library(aod)
library(car)
```

#Chunk 2: Working directory 
```{r}
getwd()
```

#Chunk 3: Read in the data 
```{r}

heart_dataframe <- read.csv("./data/heart.csv")
head(heart_dataframe)
str(heart_dataframe)
```


#Chunk 4: Data Clea-up - Check NAs and for select variable check for missing/invalid data 
Missing data checked for variables Age, RestingBP, Cholesterol and MaxHR, as these are the variables where a value zero is not valid. Missing/Zero values replaced by NA
```{r}
colSums(is.na(heart_dataframe))

invalid_Age <- heart_dataframe$Age == 0
heart_dataframe$Age[invalid_Age] <- NA

invalid_RestingBP<- heart_dataframe$RestingBP == 0
heart_dataframe$RestingBP[invalid_RestingBP] <- 132.5

#As variable cholesterol has large number(172) of invalid values. Replace zero values with random values between 1st Quartile and mean. 
invalid_Cholesterol <- heart_dataframe$Cholesterol == 0
heart_dataframe$Cholesterol[invalid_Cholesterol]<- runif(172, min = 173.2, max = 198.8)

invalid_MaxHR <- heart_dataframe$MaxHR == 0
heart_dataframe$MaxHR[invalid_MaxHR] <- NA

nrow(heart_dataframe)
colSums(is.na(heart_dataframe))
summary(heart_dataframe)
head(heart_dataframe, 50)
```

#Chunk 5: DataType Conversion. Applicable variables to factors 
```{r}
#sex
heart_dataframe$Sex[heart_dataframe$Sex == "F"] <- "Female"
heart_dataframe$Sex[heart_dataframe$Sex == "M"] <- "Male"
heart_dataframe$Sex <- as.factor(heart_dataframe$Sex)


#ChestPainType
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "TA"] <- "Typical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ATA"] <- "Atypical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "NAP"] <- "Non-anginal pain"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ASY"] <- "Asymptomatic angina"
heart_dataframe$ChestPainType <- as.factor(heart_dataframe$ChestPainType)


#FastingBS 
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 0] <- "Low  (<120 mg/dl)"
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 1] <- "High (>120 mg/dl)"
heart_dataframe$FastingBS <- as.factor(heart_dataframe$FastingBS)

#RestingECG
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "Normal"] <- "0 (normal)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "ST"] <- "1 (ST)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "LVH"] <- "2 (LVH)"
heart_dataframe$RestingECG <- as.factor(heart_dataframe$RestingECG)

#Exercise induced angina
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "N"] <- "No"
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "Y"] <- "Yes"
heart_dataframe$ExerciseAngina <- as.factor(heart_dataframe$ExerciseAngina)

#ST Slope
heart_dataframe$ST_Slope <- as.factor(heart_dataframe$ST_Slop)

#Heart Disease
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "0"] <- "Health"
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "1"] <- "Heart Disease"
heart_dataframe$HeartDisease <- as.factor(heart_dataframe$HeartDisease)

head(heart_dataframe$HeartDisease,20)
summary(heart_dataframe)
str(heart_dataframe)
```


#Chunk 6: Characterizing distributions of NUMERIC variables and correcting for skew
```{r}
#Histograms for continuous integer/dbl variables 

#AGE
age_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Age, fill = heart_dataframe$Age, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "Chocolate")+
  xlab("Age in years")+
  ylab("frequency")+
  ggtitle("Distribution of Age in database")
age_distribution 

qqplot_Age<- qplot(sample = heart_dataframe$Age)+
xlab("Age in years")+
ylab("frequency")
qqplot_Age
#qqplot o/p indicates age of subjects are normally distributed.


#CHOLESTEROL 
cholesterol_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Cholesterol, fill = heart_dataframe$Cholesterol, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "azure2")+
  xlab("Serum Cholesterol Levels in mg/dl")+
  ylab("frequency")+
  ggtitle("Distribution of Serum Cholesterol Levels across database")
cholesterol_distribution 

qqplot_cholesterol<- qplot(sample = heart_dataframe$Cholesterol, na.rm = TRUE)+
xlab("Serum Cholesterol Levels in mg/dl")+
ylab("frequency")
qqplot_cholesterol

describe(heart_dataframe$Cholesterol)
stat.desc(heart_dataframe$Cholesterol, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$Cholesterol)
#Based on qqplot, descriptive statistics and histogram, Cholesterol Levels appears to be NOT normally distributed, has a positive skew. Kurtosis of 4.6 (>3) indicates data is leptokurtic.

#log transformation to correct positive skew 
log_cholesterol <- log(heart_dataframe$Cholesterol)
log_RBP_hist <- hist(log_cholesterol, col = "azure2", border = "black", main = "Distribution of Serum Cholesterol Levels(transformed)", xlab = "Serum Cholesterol Levels in mg/dl", ylab = "frequency")

#RESTING BLOOD PRESSURE 
RBP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingBP, fill = heart_dataframe$RestingBP, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "brown")+
  xlab("Blood Pressure is mmHg")+
  ylab("frequency")+
  ggtitle("Distribution of Resting Blood Pressure Readings")
RBP_distribution 

describe(heart_dataframe$RestingBP)
stat.desc(heart_dataframe$RestingBP, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$RestingBP)

#log transformation to correct skew
log_RBP <- log(heart_dataframe$RestingBP)
log_RBP_hist <- hist(log_RBP, col = "brown", border = "black", main = "Distribution of Resting Blood Pressure Readings(transformed)", xlab = "Resting Blood Pressure in mmHg", ylab = "frequency")

#MAXIMUM HEART RATE 
MHR_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$MaxHR, fill = heart_dataframe$MaxHR))+
  geom_histogram(color = "Black", fill = "darkorchid1")+
  xlab("Maximum Heart Rate in bpm")+
  ylab("frequency")+
  ggtitle("Distribution of Maximum Heart Rate across database")
MHR_distribution 

#Distribution appears to be nearly normally distributed. 
describe(heart_dataframe$MaxHR)
stat.desc(heart_dataframe$MaxHR, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$MaxHR)


#OLD PEAK 
Oldpeak_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Oldpeak, fill = heart_dataframe$Oldpeak))+
  geom_histogram(color = "Black", fill = "cadetblue")+
  xlab("Oldpeak - ST depression induced by exercise relative to rest")+
  ylab("frequency")+
  ggtitle("Distribution of ST depression induced by exercise relative to rest")
Oldpeak_distribution #Positively skewed data 

qqplot_oldpeak<- qplot(sample = heart_dataframe$Oldpeak, na.rm = TRUE)+
xlab("Oldpeak - ST depression induced by exercise relative to rest")+
ylab("frequency")
qqplot_oldpeak

describe(heart_dataframe$Oldpeak)
stat.desc(heart_dataframe$Oldpeak, basic = FALSE, norm = TRUE)
shapiro.test(heart_dataframe$Oldpeak)  #confirms data is not normally distributed. 

#log to correct for severe positive skew 
log_oldpeak <- log(heart_dataframe$Oldpeak)
log_oldpeak
log_oldpeak_hist <- hist(log_oldpeak, col = "cadetblue", border = "black", main = "Distribution of ST depression induced by exercise relative to rest(transformed)", xlab = "Oldpeak - ST depression induced by exercise relative to rest", ylab = "frequency")


```

#Chunk 7: Characterizing distributions of FACTOR variables   
Bar plots used for factor variables and Histograms for continuous variables. 
```{r}
#SEX
gender_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Sex, fill = heart_dataframe$Sex))+
  geom_bar()+
  xlab("Gender")+
  ylab("frequency")+
  ggtitle("Distribution of Gender")+
  scale_fill_discrete(name = "Gender", labels = c("Female","Male"))
gender_distribution

#Data set contains significantly higher number of males(700) compared to females (200)

#CHEST PAIN TYPES 
CP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ChestPainType, fill = heart_dataframe$ChestPainType))+
  geom_bar()+
  xlab("Types of Chest Pain")+
  ylab("frequency")+
  ggtitle("Distribution of Types of Chest Pain")+
  scale_fill_discrete(name = "Types of Chest Pain", labels = c("Asymptomatic", "Atypica", "Non-Anginal", "Typical"))
CP_distribution

# percent distribution of chest pain types 
cp_table <- table(heart_dataframe$ChestPainType)
percent_cp <- round(cp_table/sum(cp_table) *100)
labels1 <- paste(names(cp_table), percent_cp)
labels2 <- paste(labels1, "%", sep = "")
pie(cp_table, labels = labels2, col = c("azure1", "azure2", "azure3", "azure4"), main = "Distribution of types of chest pain within heart dataset" ) 

#Atypical Angina is the most observed 

#FASTING BLOOD SUGAR LEVLS
FBS_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$FastingBS, fill = heart_dataframe$FastingBS))+
  geom_bar()+
  xlab("Fasting Blood Sugar Levels in mg/dl")+
  ylab("frequency")+
  ggtitle("Distribution of Fasting Blood Sugar Levels")+
  scale_fill_discrete(name = "Blood Sugar Levels", labels = c(">120 mg/dl","<120 mg/dl"))
FBS_distribution

#RESTING ECG
ECG_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingECG, fill = heart_dataframe$RestingECG))+
  geom_bar()+
  xlab("Resting ECG Results")+
  ylab("frequency")+
  ggtitle("Distribution of Resting ECG Results")+
  scale_fill_discrete(name = "ECG Results", labels = c("Normal", "ST", "LVH"))
ECG_distribution


#EXERCISE INDUCED ANGINA
Exangina_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ExerciseAngina, fill = heart_dataframe$ExerciseAngina))+
  geom_bar()+
  xlab("presence or absence of exercise induced angina ")+
  ylab("frequency")+
  ggtitle("Presence or Absence of Exercise Induced Angina")+
  scale_fill_discrete(name = "Exercise Induced Angina", labels = c("Absence","Presence"))
Exangina_distribution

#ST Slope
ST_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ST_Slope, fill = heart_dataframe$ST_Slope))+
  geom_bar()+
  xlab("ST Slope nature")+
  ylab("frequency")+
  ggtitle("Distribution of nature of the ST Slopes")+
  scale_fill_discrete(name = "ST Slopes", labels = c("Down", "Flat", "Up"))
ST_distribution

#HEART DISEASE
HD_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$HeartDisease, fill = heart_dataframe$HeartDisease))+
  geom_bar()+
  xlab("Heart Disease")+
  ylab("frequency")+
  ggtitle("Presence and absence of heart disease")+
  scale_fill_discrete(name = "Heart Disease", labels = c("Absence", "presence"))
HD_distribution
prop.table(table(heart_dataframe$HeartDisease)) 

#Within the data set ~55.3% have heart disease and ~44.7% do not.
```

#Chunk 8: BIAS ANALYSIS 
```{r}
# Bias analysis is conducted in order to estimate the potential magnitude and direction of biases, and to quantify the uncertainty about these biases within training datasets. For our dataset, we applied class imbalance and DPL bias analysis. Class Imbalance bias occurs when facet value d (males) has fewer training samples compared to facet a (females). We used class imbalance to assess our gender distribution. Which we have already noted, features more males than females. After performing Class Imbalance Bias Analysis, a negative CI value was found. This indicates that facet d has more training samples in the dataset. In this case we did find a negative class imbalance value. 
#bias analysis for gender distribution # Sampling Bias 
#bias analysis for gender distribution #Sampling Bias 
table(heart_dataframe["Sex"])
CI_gender <- (193-725)/(193+725)
CI_gender #-0.58 


#bias analysis - presence or absence of heart disease between gender groups (?)  
#DPL = (qa - qd) 
table(heart_dataframe$Sex, heart_dataframe$HeartDisease)

qa <- (50/193)
qd <- (458/725)
qa
qd
DPL <- (qa-qd)
DPL # -0.373 
```

<<<<<<< HEAD
#Chunk 9: Data conversion to all numeric (except HeartDisease) to obtain a complete correlation matrix and correlation plot 
=======
```{r}
#Bias Analysis - DPL 
# Difference in Proportions of Labels or DPL compares the proportion of observed positive outcomes for facet d in proportion to the observed positive outcomes of facet a. A negative DPL value indicates that facet d has a higher proportion of positive outcomes when compared with facet a. We also found DPL to reflect a negative value. 
qa <- (50/193)
qd <- (458/725)
qa
qd
DPL <- (qa-qd)
DPL # -0.373 
```

#Chunk 8: Data conversion to all numeric (except HeartDisease) to obtain a complete correlation matrix and correlation plot 
```{r}
heart_dfnum <- subset(heart_dataframe, select = -c(HeartDisease))

#YN_HeartDisease, YN_HeartDisease_bin
#ST Slope
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "2"] <- "0"
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "1"] <- "1"
heart_dfnum$ST_Slope[heart_dfnum$ST_Slope == "3"] <- "2"
heart_dfnum$ST_Slope <- as.numeric(heart_dfnum$ST_Slope)


#sex
heart_dfnum$Sex[heart_dfnum$Sex == "M"] <- "0"
heart_dfnum$Sex[heart_dfnum$Sex == "F"] <- "1"
heart_dfnum$Sex <- as.numeric (heart_dfnum$Sex)


#ChestPainType
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "TA"] <- "0"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "ATA"] <- "1"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "NAP"] <- "2"
heart_dfnum$ChestPainType[heart_dfnum$ChestPainType == "ASY"] <- "3"
heart_dfnum$ChestPainType <- as.numeric(heart_dfnum$ChestPainType)


heart_dfnum$FastingBS <- as.numeric(heart_dfnum$FastingBS)

#RestingECG
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "Normal"] <- "0"
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "ST"] <- "1"
heart_dfnum$RestingECG[heart_dfnum$RestingECG == "LVH"] <- "2"
heart_dfnum$RestingECG <- as.numeric(heart_dfnum$RestingECG)

#Exercise induced angina
heart_dfnum$ExerciseAngina[heart_dfnum$ExerciseAngina == "N"] <-"0"
heart_dfnum$ExerciseAngina[heart_dfnum$ExerciseAngina == "Y"] <- "1"
heart_dfnum$ExerciseAngina <- as.numeric(heart_dfnum$ExerciseAngina)


log_dataframe <- data.frame(heart_dataframe$Age, log_cholesterol,log_RBP, log_oldpeak, heart_dataframe$MaxHR, heart_dfnum$ST_Slope, heart_dfnum$Sex, heart_dfnum$ChestPainType, heart_dfnum$FastingBS,heart_dfnum$RestingECG, heart_dfnum$ExerciseAngina)
str(log_dataframe)
colnames(log_dataframe) <- c("Age", "Cholesterol_Level", "Resting_Blood_Pressure", "Old_Peak", "Max_Heart_Rate" , "ST_slope", "Gender","Types_of_Chest_Pain", "Fasting_Blood_Sugar_Levels", "Resting_ECG", "Exercise_Induced_Angina" )

head(heart_dfnum,20)
summary(heart_dfnum)
str(heart_dfnum)
```
 
#Chunk 10: Correlation Matrix and Plots 
```{r}
#Collrelation of all variables to numeric (except Outcome Variable) 
heart_allnum_cor <- cor(select_if (heart_dfnum, is.numeric), use = "complete.obs", method = "spearman")
heart_allnum_cor 
heart_allnum_corplot <- corrplot(heart_allnum_cor, method = "circle", order = "alphabet", tl.cex = 0.7)


#correlation of Transformed dataset 
heart_logcor <- cor(log_dataframe, use = "complete.obs", method = "spearman")
heart_logcor
heart_logcorplot <- corrplot(heart_logcor, method = "square", order = "alphabet", tl.cex = 0.7)

cor.test(log_dataframe$Age, log_dataframe$Max_Heart_Rate, alternative = "two.sided", method = "spearman") #H0 true (-ve)
cor.test(log_dataframe$Fasting_Blood_Sugar_Levels, log_dataframe$Max_Heart_Rate, alternative = "two.sided", method = "spearman")# H0  true(+ve)
cor.test(log_dataframe$ST_slope, log_dataframe$Old_Peak, alternative = "two.sided", method = "spearman") #H0 true (-ve)
cor.test(log_dataframe$ST_slope, log_dataframe$Exercise_Induced_Angina, alternative = "two.sided", method = "spearman") #Ho true (-ve)

```


#Chunk 11: Graphs: Age and Lab Values VS Heart Disease 
```{r}
#Between Heart Disease and Age 

age_HD <- ggplot(log_dataframe, aes(x = Age, fill = heart_dataframe$HeartDisease, color = heart_dataframe$HeartDisease))+
          geom_histogram(binwidth = 1, color = "black") +
          facet_wrap(vars(heart_dataframe$HeartDisease))+
          xlab("Age in Years")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Distribution of Heart Disease in relation to Age")
age_HD
#Histogram plots for presence Vs absence(Health) of heart disease have different distributions. 
#1. The distribution of presence of heart disease is negatively skewed.
#2. The distribution of absence of heart disease appears to be normally distributed. 
#3. Hence we can conclude that heart disease is more prevalent in older population compared to younger. 

#Between Resting blood pressure and presence/absence of heart disease
RBP_HD <- ggplot(log_dataframe, aes( x = Resting_Blood_Pressure, fill = heart_dataframe$HeartDisease))+
          geom_histogram()+
          xlab("Resting Blood Pressure in mmhg")+
          ylab(" Heart Disease")
# There does not appear to be a very strong relationship between resting BP and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 
RBP_HD


CHL_HD <- ggplot(log_dataframe, aes(x = Cholesterol_Level, fill = heart_dataframe$HeartDisease, color = heart_dataframe$HeartDisease))+
          geom_histogram(binwidth = 1, color = "black") +
          facet_wrap(vars(heart_dataframe$HeartDisease))+
          xlab("Serum Cholesterol Levels in mg/dl")+
          ylab("Heart Disease")+
          ggtitle("Distribution of Heart Disease in relation to Serum Cholesterol Levels")
CHL_HD

#Similarly, there does not appear to be a very strong relationship between Serum Cholesterol and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 

#Relationship between max HR achieved during exercise and presence and absence of heart disease 
MHR_HD <- ggplot(log_dataframe, aes( x = Max_Heart_Rate, fill = heart_dataframe$HeartDisease))+
          geom_histogram()+
          xlab("Maximum HR during exercise in bpm")+
          ylab("Heart Disease")
MHR_HD
#Max HR is the HR per minute reached during strenuous exercise. 
#There is a strong relationship between Max HR induced during exercise and presence and absence of heart disease. Subjects without heart disease are able to achieve a higher Max HR overall compared to subjects with heart disease. 

#Between Fasting blood sugar levels and presence/absence of heart disease
FBS_HD <- ggplot(log_dataframe, aes(x = Fasting_Blood_Sugar_Levels, fill = heart_dataframe$HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Fasting Blood Sugar levles in mgdl")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Relationship between Fasting Blood Sugar Levels and Heart Disease")
FBS_HD
#Barplot above shows comparison between fasting blood sugar levels and presence or absence of heart disease. 
# Proportions in the plot indicate that there exists a relationship between fasting blood sugar levels and heart disease. 
# Fasting blood sugar levels of >120mg/dl (high) have increased rate of presence of heart disease compared to subjects with fasting blood sugar levels of <120mg/dl (low)

```


#Chunk 12: Graphs: Cardiovascular events and Heart Disease 
```{r}
#Between Exercise induced angina and presence and absence of heart disease
ExAng_HD <- ggplot(log_dataframe, aes(x = Exercise_Induced_Angina, fill = heart_dataframe$HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Presence or Absence of exercise induced angina")+
          ylab("Heart Disease")+
          ggtitle("Relationship between Exercise Induced Angina and Heart Disease")
ExAng_HD
#Barplot above shows relationship between presence or absence of exercise induced angina Vs that of heart disease
# Proportions in the plot indicate that there exists a relationship between the two. 
# Rate of presence of exercise induced angina is significantly higher in subjects with heart disease compared to those with no heart disease (i.e. Health)


#Between ST segment depression induced by exercise and Heart Disease 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 

Oldpeak_HD <- ggplot(log_dataframe, aes(x = Old_Peak, fill = heart_dataframe$HeartDisease, color = heart_dataframe$HeartDisease))+
          geom_histogram(binwidth = 0.5, color = "black") +
          facet_wrap(vars(heart_dataframe$HeartDisease))+
          xlab("ST segment depression values in mm")+
          ylab(" Heart Disease")+
          ggtitle("ST depression induced by exercise and presence and absence of heart disease")
Oldpeak_HD

#Histograms below show the relationship between distribution of ST segment depression induced by exercise and presence and absence of heart disease. 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 
#ST segment depression distribution in subjects without heart disease(Health) is positively skewed with majority subjects at 0.0 depression. 
#However, the distribution in subjects with heart disease shows significant number of subjects with ST segment depression >0.5 which is considered pathological. 

#Distribution of chest pain types in presence vs absence of heart disease

CPT_HD <- ggplot(heart_dataframe, aes( x = HeartDisease, fill = ChestPainType))+
  geom_bar(position = "dodge")+
  xlab(" Heart Disease")+
  ylab("Type of Chest Pain")+
  ggtitle("Distribution of chest pain types in presence vs absence of heart disease")
CPT_HD


```
#Chunk 13: Logistic Regression - PsurdoR initiation
```{r}
#pseudoR calculation function for later use 
logisticPseudoR2s <- function(LogModel) {
  dev <- LogModel$deviance 
  nullDev <- LogModel$null.deviance 
  modelN <-  length(LogModel$fitted.values)
  R.l <-  1 -  dev / nullDev
  R.cs <- 1- exp ( -(nullDev - dev) / modelN)
  R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
  cat("Pseudo R^2 for logistic regression\n")
  cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
  cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
  cat("Nagelkerke R^2           ", round(R.n, 3),    "\n")
}
```



#Chunk 14: Logistic Regression
```{r}
summary(heart_dataframe)
table(heart_dataframe$HeartDisease)
prop.table(table(heart_dataframe$HeartDisease)) 
#55.34% have heart disease (508/918)
```

#Chunk 15 : Logistic Regression: Thresholding Outcome Variable HeartDisease and conversion to factor

```{r}
#assigning threshold to Heart Disease 
heart_dataframe$YN_HeartDisease[heart_dataframe$HeartDisease == "Health"] <- "0"
heart_dataframe$YN_HeartDisease[heart_dataframe$HeartDisease == "Heart Disease"] <- "1"
table(heart_dataframe$YN_HeartDisease)
str(heart_dataframe$YN_HeartDisease)

heart_dataframe$YN_HeartDisease_bin <- as.factor(heart_dataframe$YN_HeartDisease)

heart_dataframe$YN_HeartDisease <- factor(heart_dataframe$YN_HeartDisease, levels = c(0,1), labels = c("Absence", "Presence"))

table(heart_dataframe$YN_HeartDisease)
str(heart_dataframe)
```

#Chunk 16: Logistic Regression - Building Model 1 = model1_full
```{r}
str(heart_dataframe)
model1_full <- glm(YN_HeartDisease ~ Age + Sex + ChestPainType + RestingBP + Cholesterol+ FastingBS+RestingECG+ MaxHR+ ExerciseAngina+Oldpeak+ ST_Slope, data= heart_dataframe, family=binomial())
summary(model1_full)
```

#Chunk 17: Logistic Regression - Model 2 (with significant variables only )
```{r}
#model 2 - Significant variables and no interactions
model2.1_reduced <- glm(YN_HeartDisease ~ Sex + ChestPainType + FastingBS + ExerciseAngina+Oldpeak+ ST_Slope,  data = heart_dataframe, family=binomial())
model2.1_reduced

coef(model2.1_reduced)
exp(coef(model2.1_reduced))
vif(model2.1_reduced)
#Odds of Heart Disease Increases by a factor of 2.72  with every event of presence of Exercise Angina while holding rest constant.  
#Odds of Heart Disease Increased by a factor of 3.55 for every unit increase in ST_Slope(Flat) ??? while holding rest constant 
#Odds of Heart Disease increases by a factor of 1.45 for every unit of depression in ST segment induced by exercise (Oldpeak) 
#Odds of Heart Disease increases by a factor of 4.99 for subject being a Male while holding the rest constant. 



#model 2 
model2.2_reduced <- glm(YN_HeartDisease ~ Sex + ChestPainType +  ExerciseAngina+Oldpeak+ ST_Slope + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heart_dataframe, family=binomial())
model2.2_reduced

summary(model2.2_reduced) 

coef(model2.2_reduced)
exp(coef(model2.2_reduced))
vif(model2.2_reduced)

```

# Chunk 18: Removing Outliers for Model 3 
```{r}
#find outliers 
cooksd <- cooks.distance(model2.2_reduced)
cooksd
plot(cooksd, pch = "*", cex = 2) 

outlier <- c()

for (i in cooksd){
  if((is.na(i)== FALSE & i > (4/nrow(heart_dataframe))))
    {outlier <- append(outlier, i)
  }
}

length(outlier)
outlier_index <- which(cooksd %in% outlier)
outlier_index

heart_no_outlier <- heart_dataframe[-outlier_index,]
heart_no_outlier
nrow(heart_no_outlier)

```

#Chunk 19 -  Model 3.1 (model 2 with no outliers)
```{r}
head(heartdf_no_outlier)
#MODEL 3.1: model 2 with no outliers 
model3.1 <- glm(YN_HeartDisease ~ Sex + ChestPainType + Oldpeak+ ST_Slope + FastingBS + ExerciseAngina + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heartdf_no_outlier, family=binomial())


model3.1
summary(model3.1) 

coef(model3.1)
exp(coef(model3.1))
vif(model3.1)

model3.1r2  <- 1- model3.1$deviance/model3.1$null.deviance
model3.1r2

model3.1adjustedr2 <- 1-(model3.1$deviance-(ncol(heart_dataframe)-1))/model3.1$null.deviance
model3.1adjustedr2
```

#Chunk 20: MODEL 3.2:  model 3.1 without variables with high VIF 
```{r}

model3.2 <- glm(YN_HeartDisease ~  Sex + ChestPainType + ST_Slope + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heartdf_no_outlier, family=binomial())
model3.2
summary(model3.2)
   
coef(model3.2)
exp(coef(model3.2))
vif(model3.2)

model3.2r2  <- 1- model3.2$deviance/model3.2$null.deviance
model3.2r2

model3.2adjustedr2 <- 1-(model3.2$deviance-(ncol(heart_dataframe)-1))/model3.2$null.deviance
model3.2adjustedr2
```

#HYPOTHESIS TESTING - Non-parametric 

#Chunk 21: Comparing  model performance - ANOVA 
```{r}
anova(model1_full, model2.1_reduced, test = "Chisq")

#The non significant chi-square value (p = 0.22) suggests that the reduced model with all significant predictors fits as well as the full model with all predictors. 

anova(model2.1_reduced, model2.2_reduced, test = "Chisq") 

#However, the significant chi-square value(p = 0.0008) suggests that model (2.2_reduced) with predictors with high variance inflation factor removed from list of significant predictors fits better than the model(2.1_reduced) with all significant predictors. 

```
#Chunk 22: Compare models with R2 and Adjusted r2 
```{r}
R_square  <- 1- model2.2_reduced$deviance/model2.2_reduced$null.deviance
R_square

adjusted_R_square <- 1-(model2.2_reduced$deviance-(ncol(heart_dataframe)-1))/model2.2_reduced$null.deviance
adjusted_R_square

R_square  <- 1- model3.2$deviance/model3.2$null.deviance
R_square

adjusted_R_square <- 1-(model3.2$deviance-(ncol(heart_dataframe)-1))/model3.2$null.deviance
adjusted_R_square
```

#chunk 23  compare models - PseudoR2
```{r}

pseudo_2.2 <-  logisticPseudoR2s(model2.2_reduced) #model without significant predictors with high VIF 
pseudo_3.2 <- logisticPseudoR2s(model3.2) #model without outliers & significant predictors with high VIF 

```

#Chunk 24:  Confusion matrix and Accurancy: models 2.2_reduced and 3.2 using heart dataset with no outliers 
```{r}
#model2.2_reduced with heartdf_no_outlier
heartdf_no_outlier$predict_1 <- predict(model2.2_reduced, newdata=heartdf_no_outlier, type="response")

heartdf_no_outlier$predicted_YN_HeartDisease_1 <- ifelse(heartdf_no_outlier$predict_1 > 0.5,1,0)

heartdf_no_outlier$eval_1[heartdf_no_outlier$predicted_YN_HeartDisease_1==1 & heartdf_no_outlier$YN_HeartDisease_bin ==1]="True Positive"
heartdf_no_outlier$eval_1[heartdf_no_outlier$predicted_YN_HeartDisease_1==1 & heartdf_no_outlier$YN_HeartDisease_bin ==0]="False Positive"
heartdf_no_outlier$eval_1[heartdf_no_outlier$predicted_YN_HeartDisease_1==0 & heartdf_no_outlier$YN_HeartDisease_bin ==1]="False Negative"
heartdf_no_outlier$eval_1[heartdf_no_outlier$predicted_YN_HeartDisease_1==0 & heartdf_no_outlier$YN_HeartDisease_bin ==0]="True Negative"

tab_2.2_reduced <- table(heartdf_no_outlier$eval_1) #model2.2_reduced with heartdf_no_outlier
tab_2.2_reduced
accuracy_2.2 <- (337+452)/(22+37+336+450)
accuracy_2.2

#model3.2 with heartdf_no_outlier 
heartdf_no_outlier$predict_2 <- predict(model3.2, newdata=heartdf_no_outlier, type="response")

heartdf_no_outlier$predicted_YN_HeartDisease_2 <- ifelse(heartdf_no_outlier$predict_2 > 0.5,1,0)

heartdf_no_outlier$eval_2[heartdf_no_outlier$predicted_YN_HeartDisease_2==1 & heartdf_no_outlier$YN_HeartDisease_bin ==1]="True Positive"
heartdf_no_outlier$eval_2[heartdf_no_outlier$predicted_YN_HeartDisease_2==1 & heartdf_no_outlier$YN_HeartDisease_bin ==0]="False Positive"
heartdf_no_outlier$eval_2[heartdf_no_outlier$predicted_YN_HeartDisease_2==0 & heartdf_no_outlier$YN_HeartDisease_bin ==1]="False Negative"
heartdf_no_outlier$eval_2[heartdf_no_outlier$predicted_YN_HeartDisease_2==0 & heartdf_no_outlier$YN_HeartDisease_bin ==0]="True Negative"

tab_3.2 <- table(heartdf_no_outlier$eval_2)#model3.2 with heartdf_no_outlier 
tab_3.2
accuracy_3.2 <- (336+451)/(21+37+336+451)
accuracy_3.2*100


```

#Chunk 25:  ROC Curve & AUC
```{r}
ROC_curve_model2.2 <- rocit(score=heartdf_no_outlier$predicted_YN_HeartDisease_1,class=heartdf_no_outlier$YN_HeartDisease_bin)
plot(ROC_curve_model2.2)

test_ROC_model2.2 = roc(heartdf_no_outlier$YN_HeartDisease_bin ~ heartdf_no_outlier$predict_1 , plot = TRUE, print.auc = TRUE)

ROC_curve_model3.2 <- rocit(score=heartdf_no_outlier$predicted_YN_HeartDisease_2,class=heartdf_no_outlier$YN_HeartDisease_bin)
plot(ROC_curve_model3.2)

test_ROC_model3.2 = roc(heartdf_no_outlier$YN_HeartDisease_bin ~ heartdf_no_outlier$predict_2 , plot = TRUE, print.auc = TRUE)
```

#Chunk 26: Overdispersion Check. 
```{r}
dispersion_model3.2 <- deviance(model3.2)/df.residual(model3.2)
dispersion_model3.2 #0.42

#Absence of significant overdispersion. 
```

#Chunk 27: Odds Ratio and Confidence Interval 
```{r}
model3.2 <- glm(YN_HeartDisease ~ Sex +  ChestPainType + Oldpeak+ ST_Slope + FastingBS:MaxHR + ST_Slope:Oldpeak + MaxHR:Age + ST_Slope:ExerciseAngina, data = heartdf_no_outlier, family=binomial())
model3.2
summary(model3.2)
coef(model3.2)
exp(coef(model3.2))
vif(model3.2)
confint(model3.2)

#combining odds ratio and 97.5% CI 
exp(cbind(OR = coef(model3.2), confint(model3.2)))

```

#CONCLUSION
Reject Null Hypothesis and accept the alternate hypothesis: 

For every unit change in interaction between Presence of Exercise Induced Angina and Nature of the ST Slope on ECG the odds of presence of heart disease are multiplied by a factor of 2.7 while holding the other variables constant . #TEST 




