---
title: "Heart Disease"
author: "Shannon Walsh"
date: "11/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chunk 1: #Libraries 
```{r}
library(plotly)
library(stats)
library(tidyr)
library(ggplot2)
library(dplyr)
library(data.table)
library(tidyverse)
library(pastecs)
library(dbplyr)
library(Hmisc)
library(scales)
library(ROSE)
library(corrplot)
library(pscl)
library(DT)
library(caret)
library(varImp)
library(gridBase)
library(gridExtra)
```

#Chunk 2: Working directory 
```{r}
getwd()
```

#Chunk 3: Read in the data 
```{r}
heart_dataframe <- read.csv("./data/heart.csv")
head(heart_dataframe)
str(heart_dataframe)
```


#Chunk 4: Check NAs and for select variable check for missing/invalid data 
Missing data checked for variables Age, RestingBP, Cholesterol and MaxHR, as these are the variables where a value zero is not valid. Missing/Zero values replaced by NA
```{r}
colSums(is.na(heart_dataframe))

invalid_Age <- heart_dataframe$Age == 0
heart_dataframe$Age[invalid_Age] <- NA

invalid_RestingBP<- heart_dataframe$RestingBP == 0
heart_dataframe$RestingBP[invalid_RestingBP] <- NA

invalid_Cholesterol <- heart_dataframe$Cholesterol == 0
heart_dataframe$Cholesterol[invalid_Cholesterol]<- NA

invalid_MaxHR <- heart_dataframe$MaxHR == 0
heart_dataframe$MaxHR[invalid_MaxHR] <- NA

na.omit(heart_dataframe)
colSums(is.na(heart_dataframe))
summary(heart_dataframe)
head(heart_dataframe, 20)
```

#Chunk 5: DataType Conversion
```{r}
#sex
heart_dataframe$Sex[heart_dataframe$Sex == "F"] <- "Female"
heart_dataframe$Sex[heart_dataframe$Sex == "M"] <- "Male"
heart_dataframe$Sex <- as.factor(heart_dataframe$Sex)


#ChestPainType
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "TA"] <- "Typical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ATA"] <- "Atypical angina"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "NAP"] <- "Non-anginal pain"
heart_dataframe$ChestPainType[heart_dataframe$ChestPainType == "ASY"] <- "Asymptomatic angina"
heart_dataframe$ChestPainType <- as.factor(heart_dataframe$ChestPainType)


#FastingBS 
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 0] <- "Low  (<120 mg/dl)"
heart_dataframe$FastingBS[heart_dataframe$FastingBS == 1] <- "High (>120 mg/dl)"
heart_dataframe$FastingBS <- as.factor(heart_dataframe$FastingBS)

#RestingECG
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "Normal"] <- "0 (normal)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "ST"] <- "1 (ST)"
heart_dataframe$RestingECG[heart_dataframe$RestingECG == "LVH"] <- "2 (LVH)"
heart_dataframe$RestingECG <- as.factor(heart_dataframe$RestingECG)

#Exercise induced angina
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "N"] <- "No"
heart_dataframe$ExerciseAngina[heart_dataframe$ExerciseAngina == "Y"] <- "Yes"
heart_dataframe$ExerciseAngina <- as.factor(heart_dataframe$ExerciseAngina)

#ST Slope
heart_dataframe$ST_Slope <- as.factor(heart_dataframe$ST_Slop)

#Heart Disease
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "0"] <- "Health"
heart_dataframe$HeartDisease[heart_dataframe$HeartDisease == "1"] <- "Heart Disease"
heart_dataframe$HeartDisease <- as.factor(heart_dataframe$HeartDisease)

head(heart_dataframe,20)
summary(heart_dataframe)
str(heart_dataframe)
```


#Chunk 6: Characterizing distributions of variables
Bar plots used for factor variables and Histograms for continuous variables. 
```{r}

#Age
qqplot_Age<- qplot(sample = heart_dataframe$Age)+
xlab("Age in years")+
ylab("count")
qqplot_Age

#qqplot o/p indicates age of subjects are normally distributed.

barplot_age <-heart_dataframe %>%
  group_by(Age) %>%
  count() %>%
  filter(n>10) %>%
  ggplot()+
  geom_col(aes(Age,n), fill = "Blue")+
  ggtitle("Age distribution of the Heart Disease Dataset")
barplot_age
#Age of majority of subjects falls between 50-60years. Plot shows a very slight negative skew. 

describe (heart_dataframe$Age)
#Mean age 54yrs 

stat.desc(heart_dataframe$Age, basic = FALSE, norm = TRUE)

#Sex
gender_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Sex, fill = heart_dataframe$Sex))+
  geom_bar()+
  xlab("Gender")+
  ylab("count")+
  ggtitle("Distribution of Gender")+
  scale_fill_discrete(name = "Gender", labels = c("Female","Male"))
gender_distribution
#Data set contains significantly higher number of males(700) compared to females (200)

#Chest Pain Types
CP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ChestPainType, fill = heart_dataframe$ChestPainType))+
  geom_bar()+
  xlab("Types of Chest Pain")+
  ylab("Count")+
  ggtitle("Distribution of Types of Chest Pain")+
  scale_fill_discrete(name = "Types of Chest Pain", labels = c("Asymptomatic", "Atypica", "Non-Anginal", "Typical"))
CP_distribution

# percent distribution of chest pain types 
cp_table <- table(heart_dataframe$ChestPainType)
percent_cp <- round(cp_table/sum(cp_table) *100)
labels1 <- paste(names(cp_table), percent_cp)
labels2 <- paste(labels1, "%", sep = "")
pie(cp_table, labels = labels2, col = c("azure1", "azure2", "azure3", "azure4"), main = "Distribution of types of chest pain within heart dataset" ) 

#Atypical Angina is the most observed 

#Fasting blood sugar 
BS_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$FastingBS, fill = heart_dataframe$FastingBS))+
  geom_bar()+
  xlab("Fasting Blood Sugar Levels in mg/dl")+
  ylab("count")+
  ggtitle("Distribution of Fasting Blood Sugar Levels")+
  scale_fill_discrete(name = "Blood Sugar Levels", labels = c(">120 mg/dl","<120 mg/dl"))
BS_distribution

#Resting ECG
ECG_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingECG, fill = heart_dataframe$RestingECG))+
  geom_bar()+
  xlab("Resting ECG Results")+
  ylab("count")+
  ggtitle("Distribution of Resting ECG Results")+
  scale_fill_discrete(name = "ECG Results", labels = c("Normal", "ST", "LVH"))
ECG_distribution


#Exercise induced angina
Exangina_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ExerciseAngina, fill = heart_dataframe$ExerciseAngina))+
  geom_bar()+
  xlab("presence or absence of exercise induced angina ")+
  ylab("count")+
  ggtitle("Presence or Absence of Exercise Induced Angina")+
  scale_fill_discrete(name = "Exercise Induced Angina", labels = c("Absence","Presence"))
Exangina_distribution

#ST Slope
ST_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$ST_Slope, fill = heart_dataframe$ST_Slope))+
  geom_bar()+
  xlab("ST Slope nature")+
  ylab("count")+
  ggtitle("Distribution of nature of the ST Slopes")+
  scale_fill_discrete(name = "ST Slopes", labels = c("Down", "Flat", "Up"))
ST_distribution

#Heart disease 
heartdisease_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$HeartDisease, fill = heart_dataframe$HeartDisease))+
  geom_bar()+
  xlab("Heart Disease")+
  ylab("count")+
  ggtitle("Presence and absence of heart disease")+
  scale_fill_discrete(name = "Heart Disease", labels = c("Absence", "presence"))
heartdisease_distribution
prop.table(table(heart_dataframe$HeartDisease)) 

#Within the data set ~55.3% have heart disease and ~44.7% do not. Which is nearly a balanced dataset. 


#Histograms for continuous integer/dbl variables 
#Cholesterol
cholesterol_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Cholesterol, fill = heart_dataframe$Cholesterol, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "Chocolate")+
  xlab("Serum Cholesterol Levels in mg/dl")+
  ylab("count")+
  ggtitle("Distribution of Serum Cholesterol Levels across database")
cholesterol_distribution 

qqplot_cholesterol<- qplot(sample = heart_dataframe$Cholesterol, na.rm = TRUE)+
xlab("Serum Cholesterol Levels in mg/dl")+
ylab("count")
qqplot_cholesterol
describe(heart_dataframe$Cholesterol)
stat.desc(heart_dataframe$Cholesterol, basic = FALSE, norm = TRUE)

#Cholesterol Levels appears to be not normally distributed, has a positive skew. The distribution also shows significant outliers to the right which may make the distribution to the positively skewed. Kurtosis of 4.47 (>3) indicates data is leptokurtic(heavy tails)


#Resting systolic BP 
BP_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$RestingBP, fill = heart_dataframe$RestingBP, na.rm = TRUE))+
  geom_histogram(color = "Black", fill = "brown")+
  xlab("Blood Pressure is mmHg")+
  ylab("count")+
  ggtitle("Distribution of Resting Blood Pressure Readings")
BP_distribution 

describe(heart_dataframe$RestingBP)
stat.desc(heart_dataframe$RestingBP, basic = FALSE, norm = TRUE)
#Mean SBP  132.5mmHg


#MaxHR
HR_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$MaxHR, fill = heart_dataframe$MaxHR))+
  geom_histogram(color = "Black", fill = "darkorchid1")+
  xlab("Maximum Heart Rate in bpm")+
  ylab("count")+
  ggtitle("Distribution of Maximum Heart Rate across database")
HR_distribution 
stat.desc(heart_dataframe$MaxHR, basic = FALSE, norm = TRUE)

#Data is slightly Negatively skewed. Majority of the subjects' MaxHR falls 125-150bpm and data is Leptokurtik with kurtosis of -4.6 i.e. Heavy tailed with values father from mean.  

#oldpeak
Oldpeak_distribution <- ggplot(heart_dataframe, aes( x= heart_dataframe$Oldpeak, fill = heart_dataframe$Oldpeak))+
  geom_histogram(color = "Black", fill = "cadetblue")+
  xlab("Oldpeak - ST depression induced by exercise relative to rest")+
  ylab("count")+
  ggtitle("Distribution of ST depression induced by exercise relative to rest")
Oldpeak_distribution 
stat.desc(heart_dataframe$Oldpeak, basic = FALSE, norm = TRUE)

#Positively skewed data 
```





#Correlation: Age and Lab Values VS Heart Disease 
```{r}
#Between Heart Disease and Age 
age_HD <- ggplot(heart_dataframe, aes(x = Age, fill = HeartDisease, color = HeartDisease))+
          geom_histogram(binwidth = 1, color = "black") +
          facet_wrap(vars(HeartDisease))+
          xlab("Age in Years")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Distribution of Heart Disease in relation to Age")
age_HD
#Histogram plots for presence Vs absence(Health) of heart disease have different distributions. 
#1. The distribution of presence of heart disease is negatively skewed.
#2. The distribution of absence of heart disease appears to be normally distributed. 
#3. Hence we can conclude that heart disease is more prevalent in older population compared to younger. 


#Between Fasting blood sugar levels and presence/absence of heart disease
FBS_HD <- ggplot(heart_dataframe, aes(x = FastingBS, fill = HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Fasting Blood Sugar levles in mgdl")+
          ylab("Frequency of presence or absence of Heart Disease")+
          ggtitle("Relationship between Fasting Blood Sugar Levels and Heart Disease")
FBS_HD
#Barplot above shows comparison between fasting blood sugar levels and presence or absence of heart disease. 
# Proportions in the plot indicate that there exists a relationship between fasting blood sugar levels and heart disease. 
# Fasting blood sugar levels of >120mg/dl (high) have increased rate of presence of heart disease compared to subjects with fasting blood sugar levels of <120mg/dl (low)

#Between Resting blood pressure and presence/absence of heart disease
RBP_HD <- ggplot(heart_dataframe, aes( x = RestingBP, y = HeartDisease, fill = HeartDisease))+
          geom_boxplot()+
          xlab("Resting Blood Pressure in mmhg")+
          ylab(" Heart Disease")
# There does not appear to be a very strong relationship between resting BP and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 

CHL_HD <- ggplot(heart_dataframe, aes( x = Cholesterol, y = HeartDisease, fill = HeartDisease))+
          geom_boxplot()+
          xlab("Serum Cholesterol Levels in mg/dl")+
          ylab("Heart Disease")
#Similarly, there does not appear to be a very strong relationship between Serum Cholesterol and Presence and absence of heart disease as the median (middle 50%) of values for both plots are very similar. 

#Relationship between max HR achieved during exercise and presence and absence of heart disease 
MHR_HD <- ggplot(heart_dataframe, aes( x = MaxHR, y = HeartDisease, fill = HeartDisease))+
          geom_boxplot()+
          xlab("Maximum HR during exercise in bpm")+
          ylab("Heart Disease")
#Max HR is the HR per minute reached during strenuous exercise. 
#There is a strong relationship between Max HR induced during exercise and presence and absence of heart disease. Subjects without heart disease are able to achieve a higher Max HR overall compared to subjects with heart disease. 

grid.arrange(RBP_HD, CHL_HD, MHR_HD)
```


#Correlation: Cardiovascular events and Heart Disease 
```{r}
#Between Exercise induced angina and presence and absence of heart disease
ExAng_HD <- FBS_HD <- ggplot(heart_dataframe, aes(x = ExerciseAngina, fill = HeartDisease))+
          geom_bar(position = "fill") +
          xlab("Presence or Absence of exercise induced angina")+
          ylab("Heart Disease")+
          ggtitle("Relationship between Exercise Induced Angina and Heart Disease")
ExAng_HD
#Barplot above shows relationship between presence or absence of exercise induced angina Vs that of heart disease
# Proportions in the plot indicate that there exists a relationship between the two. 
# Rate of presence of exercise induced angina is significantly higher in subjects with heart disease compared to those with no heart disease (i.e. Health)

#Between ST segment depression induced by exercise and Heart Disease 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 
Oldpeak_HD <- ggplot(heart_dataframe, aes(x = Oldpeak, fill = HeartDisease, color = HeartDisease))+
          geom_histogram(binwidth = 0.5, color = "black") +
          facet_wrap(vars(HeartDisease))+
          xlab("ST segment depression values in mm")+
          ylab(" Heart Disease")+
          ggtitle("ST depression induced by exercise and presence and absence of heart disease")
Oldpeak_HD
#Histrograms below show the relationship between distribution of ST segment depression induced by exercise and presence and absence of heart disease. 
#Note: ST segment depression less than 0.5 mm is accepted in all leads. ST segment depression 0.5 mm or more is considered pathological. 
#ST segment depression distribution in subjects without heart disease(Health) is positively skewed with majority subjects at 0.0 depression. 
#However, the distribution in subjects with heart disease shows significant number of subjects with ST segment depression >0.5 which is considered pathological. 

#Distribution of chest pain types in presence vs absence of heart disease
CPT_HD <- ggplot(heart_dataframe, aes( x = HeartDisease, fill = ChestPainType))+
  geom_bar(position = "dodge")+
  xlab(" Heart Disease")+
  ylab("Type of Chest Pain")+
  ggtitle("Distribution of chest pain types in presence vs absence of heart disease")
CPT_HD


```
#Correlation cont. 
```{r}

```

